# Clustering and Differential Expression {#sec-img-clustering-de}

## Preamble

### Introduction

* Clustering is an important step for the workflow analysis
* Many methods are fouscing on the development of methods based-off scRNA clusteirng method. But 
imaging-based datasets have a different strategies and cannot make the same assumption as the 
other 

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
#library(dplyr)
#library(tidyr)
library(patchwork)
library(ggspavis)
library(scran)
library(scater)
library(SpatialExperiment)
library(igraph)
library(Banksy)
```


```{r load data, message=FALSE, warning=FALSE, eval=FALSE}
seed <- 2025
# load data from previous section
spe <- readRDS("img-spe_qc.rds")
spe_CosMX <- STexampleData::CosMx_lungCancer()
# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        legend.key.size=unit(0.5, "lines"),
        panel.background=element_rect(fill="black", color="white")))
```

## Preprocessing

Library size-based normalization as is typical for scRNA-seq data has been shown to be problematic for ST data, especially so for targeted panels underlying current commercial imaging-based ST platforms [@Atta2024,@Bhuva2024].
For lack of a better approach, we here use standard log-library size normalization.
We caution readers, however, to keep an eye out in the literature for attempts to provide a better strategy.

```{r calc-norm, eval=FALSE}
spe <- scater::logNormCounts(spe)
```

## Clustering

### Non-spatial Clustering

1. Using Leiden clustering to showcase the results
2. Can also use spatialPCA or BanksyPCA results for some already spatially-aware clusterings

```{r leiden clustering, cache=TRUE, eval=FALSE}
spe <- scater::runPCA(spe, ncomponents=30) 
g <- scran::buildSNNGraph(spe, use.dimred="PCA", k=30)

set.seed(seed)
k <- igraph::cluster_leiden(g, resolution=0.05)$membership
table(spe$Leiden_clusters <- factor(k, seq_along(unique(k))))
```

### Spatially-aware clustering: Neighborhood smoothing methods

1. Brief introduction of BANKSY principle
2. Introduction of the type of methods that are embedding the spatial information
3. Suggest also other methods with similar line of logic (embedding-based)

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] essentially computes PCs that are weighted by embeds cells in a product space of their own and the local neighborhood transcriptome, representing cell state and microenvironment.
A key parameter for this method is $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), which controls the amount of spatial information to incorporate; specifically, **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r clust-sp, message=FALSE, cache=TRUE, eval=FALSE}
lambda <- 0.8
k_geom <- 30
npcs <- 50

set.seed(seed)
spe <- Banksy::computeBanksy(spe, assay_name = "logcounts", k_geom = k_geom)
spe <- Banksy::runBanksyPCA(spe, lambda = lambda, npcs = npcs)
spe <- Banksy::clusterBanksy(spe, lambda = lambda, npcs = npcs, resolution = 0.3)

k <- grep("^clust", names(colData(spe)))
names(colData(spe))[k] <- "Banksy_clusters"
table(spe$Banksy_clusters <- factor(spe$Banksy_clusters, 
                                    seq_along(unique(spe$Banksy_clusters))))
```

### Spatially-aware clustering: Encoder-based methods

1. Introduction of encoder-based methods
2. INtroduction of CellCharter methods specifically
3. Other methods: STAGATE PROST SpaceFlow

```{r}
library(basilisk)
# get basilisk
py_env <- basilisk::BasiliskEnvironment(
  envname="CellCharter_env",
  pkgname="base",
  channels = c("conda-forge", "pytorch"),
  packages=c(
    "python=3.10.15",
    "pytorch=1.12.1",
    "torchvision=0.13.1",
    "torchaudio=0.12.1"
  ),
  pip=c(
    "scvi-tools==0.20.3",
    "cellcharter==0.2.0",
    "anndata==0.10.9"
  )
)

# run basilisk
print(py_env)
proc <- basilisk::basiliskStart(py_env)
```

```{r}
library(reticulate)
reticulate::use_condaenv(basilisk::obtainEnvironmentPath(py_env))
print(reticulate::py_config())
```

```{python}
import cellcharter as cc
import scvi
import squidpy as sq
import anndata as ad
import numpy as np
import pandas as pd
import random

adata = ad.read_h5ad("spe.h5ad")
adata.layers["counts"] = adata.X
seed = 2025
random.seed(seed)
scvi.settings.seed = seed

# Variational autoencoder for feature extraction
scvi.model.SCVI.setup_anndata(adata, layer="counts")
model = scvi.model.SCVI(adata)
model.train(early_stopping=True, enable_progress_bar=True)
adata.obsm['X_scVI'] = model.get_latent_representation(adata).astype(np.float32)

# Getting neighborhood aggregation
sq.gr.spatial_neighbors(adata, library_key='sample', coord_type='generic', 
                        delaunay=True, spatial_key='spatial_fov', percentile=99)
cc.gr.aggregate_neighbors(adata, n_layers=3, use_rep='X_scVI', 
                          out_key='X_cellcharter', sample_key='sample')

# Clustering by scanning a range of data number
model = cc.tl.Cluster(n_clusters=14, random_state=seed)
model.fit(adata, use_rep='X_cellcharter')

label_df = pd.DataFrame({"label":model.predict(adata, use_rep='X_cellcharter')}) 
label_df[["label"]].value_counts()
"""
```

```{r load cellcharter results, eval=FALSE}
cc_labels <- read.csv("label_cellcharter.csv")[["label"]]
table(spe$CellCharter_clusters <- as.factor(cc_labels))
```

### Comparison and evaluation of methods

* Visualization (Align the cluster labels before running it)
* Smoothness entropy checking

```{r plot-xy-k-joint, fig.width=8, fig.height=3, eval=FALSE}
#| collapse: true
df <- data.frame(colData(spe), spatialCoords(spe))
nk <- max(
    length(unique(df$Leiden_clusters)), 
    length(unique(df$Banksy_clusters)),
    length(unique(df$CellCharter_clusters)))
pal <- hcl.colors(nk, "Spectral")
ggplot(df, 
    aes(x_centroid, y_centroid, col=Leiden_clusters)) +
    ggtitle("SNN graph-based Louvain clustering") +
ggplot(df, 
    aes(x_centroid, y_centroid, col=Banksy_clusters)) +
    ggtitle(expression("Banksy clustering ("*lambda~"0.2)")) +
ggplot(df, 
    aes(x_centroid, y_centroid, col=CellCharter_clusters)) +
    ggtitle("CellCharter clustering") +
plot_layout(nrow=1) &
    geom_point(shape=16, stroke=0, size=0.1) &
    scale_color_manual(values=pal) & theme_xy &
    guides(col=guide_legend(override.aes=list(size=2))) & 
    theme(plot.title = element_text(size=8),
          legend.title=element_text(size=5),
          legend.text=element_text(size=5))
```

Especially in large tissues, the above type of plot makes it difficult to spot rare subpopulations, and might cause cells to overlap in regions with high cellular density.
This can be misleading, as we will tend to see only highly abundant subpopulations, or the cells plotted last and on top (i.e., later columns in the object).

To better distinguish between different subpopulations, we can instead generate separate spatial plots with one subpopulation highlighted at a time:

```{r plot-xy-k-split, collapse=TRUE, fig.width=12, fig.height=4, eval=FALSE}
ps <- lapply(levels(df$Banksy_clusters), \(.) 
    ggplot(df, aes(x_centroid, y_centroid, col=(Banksy_clusters == .))) + 
    scale_color_manual(values=c("navy", "cyan")) + ggtitle(.))
wrap_plots(ps, nrow=2) & 
    coord_equal(expand=FALSE) &
    geom_point(shape=16, stroke=0, size=0.2) &
    theme_void() & theme(
        legend.position="none",
        plot.title=element_text(hjust=0.5),
        panel.background=element_rect(fill="black", color="white"))
```


### Other spatially-aware clustering methods: Probability statistics-based methods 

* See more example in @sec-seq-clustering-de
* BASS
* DR.SC

### (Collapse) Effect of configurations on the results

* Cross-methods ARI validation

### (Collapse) Scalability of clusteirng methods on the results

* Using local machine to show the runtime availablility 
* Refer to benchmarking studies for potential scability issue

## Differential Expression Analysis

For details on identifying genes that are differentially expressed (DE) between groups of cells, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/marker-detection.html); a standard approach is given below, visualizing the average expression of exemplary markers across clusters:

```{r DEG analysis, eval=FALSE, fig.width=8, fig.height=4}
#| code-fold: true
#| label: fig-charts
#| fig-cap: "Charts"
#| fig-subcap: 
#|   - "First"
#|   - "Second"
#| layout-ncol: 2

getTopDEGs <- function(spe, clustering){
  mgs <- scran::findMarkers(spe, groups=spe[[clustering]], direction="up")
  top <- lapply(mgs, \(.) rownames(.)[.$Top <= 1])
  top <- unique(unlist(top))
  top <- top[top %in% rownames(assays(spe)[["logcounts"]])]
  
  pbs <- aggregateAcrossCells(spe, 
    ids=spe[[clustering]], subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
  
  return(pbs)
}
pbsLeiden <- getTopDEGs(spe = spe, clustering = "Leiden_clusters")
pbsBanksy <- getTopDEGs(spe = spe, clustering = "Banksy_clusters")

scater::plotHeatmap(pbsLeiden, 
                    colour_columns_by="Leiden_clusters", fontsize=9,
                    features=rownames(pbsLeiden), center=TRUE, scale=TRUE,
                    cluster_cols=FALSE, cluster_rows=FALSE)

scater::plotHeatmap(pbsBanksy, 
                    colour_columns_by="Banksy_clusters", fontsize=9,
                    features=rownames(pbsBanksy), center=TRUE, scale=TRUE,
                    cluster_cols=FALSE, cluster_rows=FALSE)
```

```{r plot-xy-expr, fig.width=4, fig.height=3, eval=FALSE, message=FALSE}
#| collapse: TRUE
es <- as.matrix(t(logcounts(spe)[rownames(pbsBanksy), ]))
fd <- data.frame(df, es, check.names=FALSE)
gene_plot <- ggplot(fd, aes(x_centroid, y_centroid, col=MZB1)) +
    coord_equal(expand=FALSE) +
    geom_point(shape=16, stroke=0, size=0.2) +
    scale_color_gradientn(colors=c("navy", "red", "gold", "ivory")) +
    theme_xy + theme(legend.key.height=unit(1, "lines"))

cluster_plot <- ggplot(df, aes(x_centroid, y_centroid, col=(Banksy_clusters ==3))) + 
  geom_point(shape=16, stroke=0, size=0.2) +
  scale_color_manual(values=c("navy", "cyan")) + 
  ggtitle("Banksy Cluster 3") + 
  coord_equal(expand=FALSE) +
  theme_void() + 
  theme(legend.position="none",
        plot.title=element_text(hjust=0.5),
        panel.background=element_rect(fill="black", color="white"))

gene_plot | cluster_plot
```

### Comparison of DEGs in different spatially-aware clustering cases (collapse)

## Appendix

### Saving {.unnumbered}

```{r save-data, eval=FALSE}
# default to transcription-based clustering
spe$k <- spe$Banksy_clusters; saveRDS(spe, "img-spe_cl.rds")
```

### Benchmarks {.unnumbered}

* https://doi.org/10.1093/bfgp/elae040
* https://doi.org/10.1186/s13059-024-03361-0 
* https://doi.org/10.1038/s41592-024-02325-3 
* https://doi.org/10.1016/j.csbj.2023.11.055 

### References {.unnumbered}


