# Quality control

- `r BiocStyle::Githubpkg("drighelli/SpaceTrooper")`
- data are missing molecule coordinates and polygon boundaries to due more sophisticated things...

## Preamble

### Introduction

The analyses demonstrated here will use 313-plex Xenium data (10x Genomics) of a human breast cancer biopsy section [@Janesick2023] and the 1k-plex CosMX data (Nanostring) of a human lung non-small cell lung cancer [@He2022-CosMx], both available through the `r BiocStyle::Biocpkg("STexampleData")` package as part of Bioconductor's `r BiocStyle::Biocpkg("ExperimentHub")`.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(SpatialExperiment)
library(SpaceTrooper)
library(STexampleData)
```

## Data creation (temp)

This is a temporary section to integrate the datasets in `r BiocStyle::Biocpkg("STexampleData")` with
polygon information, useful for the visualization plots presented in this chapter.
Ideally, a new version of `r BiocStyle::Biocpkg("STexampleData")` will ship the
polygons directly within the object.

### CosMX

First, we load the CosMX data from `r BiocStyle::Biocpkg("STexampleData")`:

```{r load-cosmx, message=FALSE, warning=FALSE}
cosmx <- CosMx_lungCancer()
cosmx
```

Then, we use `r BiocStyle::Biocpkg("SpaceTrooper")` to add the polygon and FOV information:

```{r load-cosmx-poly, message=FALSE, warning=FALSE}
poly <- readPolygonsCosmx("../../Lung9_Rep1-polygons.csv")
fov <- read.csv("../../Lung9_Rep1_fov_positions_file.csv")

cd <- data.frame(colData(cosmx))
poly <- left_join(poly, cd, 
            by = c("fov", "cellID"="cell_ID"))
cd <- left_join(cd, poly, 
      by = c("fov", "cell_ID"="cellID"))
cd <- dplyr::rename(cd, polygons = global)
colData(cosmx) <- as(cd, "DataFrame")
metadata(cosmx)$fov_positions <- fov
metadata(cosmx)$fov_dim = c(xdim = 5472, ydim = 3648)
```

### Xenium

We now load the Xenium data from `r BiocStyle::Biocpkg("STexampleData")`:

```{r load-xenium, message=FALSE, warning=FALSE}
xenium <- Janesick_breastCancer_Xenium_rep1()
# check object
xenium
```

Then, we use `r BiocStyle::Biocpkg("SpaceTrooper")` to add the polygon information:

```{r load-xenium-poly, message=FALSE, warning=FALSE}
poly <- readPolygonsXenium("../../cell_boundaries.parquet")

xenium <- addPolygonsToSPE(xenium, poly)
colData(xenium)
```

## Visualization

### CosMX

- Cell centroids

```{r plot-cosmx-xy, echo=FALSE, out.width="66%", fig.align="center"}
plotCentroidsSPE(cosmx)
```
- FOV map (cosmx only)

```{r plot-fov-map, echo=FALSE, out.width="66%", fig.align="center"}
plotCellsFovs(cosmx)
```

- Polygons (SpaceTrooper)

```{r plot-cosmx-poly, echo=FALSE, out.width="66%", fig.align="center"}
plotPolygonsSPE(cosmx[,cosmx$fov==9])
```

### Xenium

- Cell centroids

```{r plot-xenium-xy, echo=FALSE, out.width="66%", fig.align="center"}
plotCentroidsSPE(xenium)
```
- Polygons (SpaceTrooper)

```{r plot-xenium-poly, echo=FALSE, out.width="66%", fig.align="center"}
box <- (spatialCoords(xenium)[,1] >= 4000 & spatialCoords(xenium)[,1] <= 6000) &
  (spatialCoords(xenium)[,2] >= 0 & spatialCoords(xenium)[,2] <= 2000)
plotPolygonsSPE(xenium[,box], colour_by = "transcript_counts")
```


## QC Metrics

- Transcript based (total count, ...)
- Morphology based (Area, ..)
- Image based (DAPI, ...)
- Border effecsts (only CosMx)

Including histograms and violin plots and PCA plots etc

- Quality score (i.e., combining the metrics)

## Filtering

- Medcouple on individual metrics
- Filtering based on QS

## Appendix

### References {.unnumbered}
