# Clustering and Differential Expression {#sec-img-clustering-de}

## Preamble

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
#library(dplyr)
#library(tidyr)
#library(patchwork)
library(ggspavis)
library(scran)
library(scater)
library(SpatialExperiment)
library(igraph)
library(Banksy)
library(clue)
```


```{r load data, message=FALSE, warning=FALSE}
seed <- 2025
# load data from previous section
spe <- readRDS("img-spe_qc.rds")
spe_CosMX <- STexampleData::CosMx_lungCancer()
# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        legend.key.size=unit(0.5, "lines"),
        panel.background=element_rect(fill="black", color="white")))
```

## Preprocessing

Library size-based normalization as is typical for scRNA-seq data has been shown to be problematic for ST data, especially so for targeted panels underlying current commercial imaging-based ST platforms [@Atta2024,@Bhuva2024].
For lack of a better approach, we here use standard log-library size normalization.
We caution readers, however, to keep an eye out in the literature for attempts to provide a better strategy.

```{r calc-norm}
spe <- scater::logNormCounts(spe)
```

## Clustering

### Non-spatial Clustering

1. Using Leiden clustering to showcase the results
2. Can also use spatialPCA or BanksyPCA results for some already spatially-aware clusterings

```{r leiden clustering}
spe <- scater::runPCA(spe, ncomponents=30) 
g <- scran::buildSNNGraph(spe, use.dimred="PCA", k=30)

set.seed(seed)
k <- igraph::cluster_leiden(g, resolution=0.09)$membership
table(spe$Leiden_clusters <- factor(k, seq_along(unique(k))))
```

### Spatially-aware clustering: Neighborhood smoothing methods

1. Brief introduction of BANKSY principle
2. Introduction of the type of methods that are embedding the spatial information
3. Suggest also other methods with similar line of logic (embedding-based)

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] essentially computes PCs that are weighted by embeds cells in a product space of their own and the local neighborhood transcriptome, representing cell state and microenvironment.
A key parameter for this method is $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), which controls the amount of spatial information to incorporate; specifically, **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r clust-sp, message=FALSE, cache=TRUE}
lambda <- 0.8
k_geom <- 30
npcs <- 50

set.seed(seed)
spe <- Banksy::computeBanksy(spe, assay_name = "logcounts", k_geom = k_geom)
spe <- Banksy::runBanksyPCA(spe, lambda = lambda, npcs = npcs)
spe <- Banksy::clusterBanksy(spe, lambda = lambda, npcs = npcs, resolution = 0.5)

k <- grep("^clust", names(colData(spe)))
names(colData(spe))[k] <- "Banksy_clusters"
table(spe$Banksy_clusters <- as.factor(spe$Banksy_clusters))
```


### Spatially-aware clustering: Encoder-based methods

1. Introduction of encoder-based methods
2. INtroduction of CellCharter methods specifically
3. Other methods: STAGATE PROST SpaceFlow

```{python, }
import cellcharter as cc
import squidpy as sq
import anndata as ad
import numpy as np
import pandas as pd

#TODO add adata conversion of raw spe data
adata = 
seed = 2025
random.seed(seed)
scvi.settings.seed = seed

# Variational autoencoder for feature extraction
scvi.model.SCVI.setup_anndata(adata, layer="counts")
model = scvi.model.SCVI(adata)
model.train(early_stopping=True, enable_progress_bar=False)
adata.obsm['X_scVI'] = model.get_latent_representation(adata).astype(np.float32)

# Getting neighborhood aggregation
sq.gr.spatial_neighbors(adata, library_key='sample', coord_type='generic', 
                        delaunay=True, spatial_key='spatial_fov', percentile=99)
cc.gr.aggregate_neighbors(adata, n_layers=3, use_rep='X_scVI', 
                          out_key='X_cellcharter', sample_key='sample')

# Clustering by scanning a range of data number
model = cc.tl.Cluster(n_clusters=14, random_state=seed)
model.fit(adata, use_rep='X_cellcharter')

label_df = pd.DataFrame({"label":model.predict(adata, use_rep='X_cellcharter')}) 
label_df[["label"]].value_counts()
```

### Comparison and evaluation of methods

* Visualization (Align the cluster labels before running it)
* Smoothness entropy checking
* Some other evaluation metrics?

```{r Align all the methods}
df <- data.frame(colData(spe), spatialCoords(spe))

columns <- c("Banksy_clusters", "Leiden_clusters", "CellCharter_clusters")

for (c in columns){
    hung <- clue::solve_LSAP(table(df[,"Banksy_clusters"], df[,c]), 
                             maximum = TRUE)
    lookup <- cbind(seq_along(hung), levels(df[[c]])[hung])
    levels(df[,c]) <- lookup[order(as.integer(lookup[,2])),1]
    df[,c] <- as.factor(as.character(df[,c]))
}
```

```{r plot-xy-k-joint, fig.width=8, fig.height=3}
#| collapse: true
nk <- max(
    length(unique(df$Leiden_clusters)), 
    length(unique(df$Banksy_clusters)),
    length(unique(df$CellCharter_clusters)))
pal <- hcl.colors(nk, "Spectral")
ggplot(df, 
    aes(x_centroid, y_centroid, col=Leiden_clusters)) +
    ggtitle("SNN graph-based Louvain clustering") +
ggplot(df, 
    aes(x_centroid, y_centroid, col=Banksy_clusters)) +
    ggtitle(expression("Banksy clustering ("*lambda~"0.2)")) +
ggplot(df, 
    aes(x_centroid, y_centroid, col=CellCharter_clusters)) +
    ggtitle("CellCharter clustering") +
plot_layout(nrow=1) &
    geom_point(shape=16, stroke=0, size=0.1) &
    scale_color_manual(values=pal) & theme_xy &
    guides(col=guide_legend(override.aes=list(size=1))) & 
    theme(plot.title = element_text(size = 8))
```

### Other spatially-aware clustering methods: Bayesian-inference methods 

* BASS
* DR.SC

### (Collapse) Effect of configurations on the results

* Cross-methods ARI validation

### (Collapse) Scalability of clusteirng methods on the results


## Differential Expression Analysis

```{r DEG analysis for leiden clustering}
#| code-fold: true

getPBS <- function(spe, clustering){
  mgs <- findMarkers(spe, groups=spe[[clustering]], direction="up")
  top <- lapply(mgs, \(.) rownames(.)[.$Top <= 1])
  top <- unique(unlist(top))
  pbs <- aggregateAcrossCells(spe, 
      ids=spe[[clustering]], subset.row=top, 
      use.assay.type="logcounts", statistics="mean")
  return(pbs)
}

pbsLeiden <- getPBS(spe = spe, clustering = "Leiden_clustering")
pbsBanksy <- getPBS(spe = spe, clustering = "Banksy_clustering")

HLeiden <- plotHeatmap(pbsLeiden, 
    colour_columns_by="Leiden_clusters", fontsize=9,
    features=top, center=TRUE, scale=TRUE,
    cluster_cols=FALSE, cluster_rows=FALSE)
HBanksy <- plotHeatmap(pbsBanksy, 
    colour_columns_by="Banksy_clusters", fontsize=9,
    features=top, center=TRUE, scale=TRUE,
    cluster_cols=FALSE, cluster_rows=FALSE)

grid.arrange(HLeiden$gtable, HBanksy$gtable, ncol = 2)
```

```{r plot-xy-expr, fig.width=4, fig.height=3}
#| collapse: TRUE
es <- as.matrix(t(logcounts(spe)[top, ]))
fd <- data.frame(df, es, check.names=FALSE)
ggplot(fd, aes(x_centroid, y_centroid, col=KRT14)) +
    coord_equal(expand=FALSE) +
    geom_point(shape=16, stroke=0, size=0.2) +
    scale_color_gradientn(colors=c("navy", "red", "gold", "ivory")) +
    theme_xy + theme(legend.key.height=unit(1, "lines"))
```

### Comparison of DEGs in different spatially-aware clustering cases (collapse)

## Appendix

### Existing Benchmark

* https://doi.org/10.1093/bfgp/elae040
* https://doi.org/10.1186/s13059-024-03361-0 
* https://doi.org/10.1038/s41592-024-02325-3 
* https://doi.org/10.1016/j.csbj.2023.11.055 

### Saving data

```{r save-data}
# default to transcription-based clustering
spe$k <- spe$k_tx; saveRDS(spe, "img-spe_cl.rds")
```

### Session info

## References {.unnumbered}
