# Workflow: Visium CRC {#sec-seq-workflow-visium-crc}

## Preamble

### Introduction

In this demo, we will be analyzing Visium data on a human colorectal cancer biopsy from @Oliveira2024 (@sec-bkg-datasets). Rather than recapitulating all possible analyses, our goal is to highlight those that might be of particular interest in the context of these data.

### Dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(osfr)
library(scran)
library(scater)
library(igraph)
library(AUCell)
library(scuttle)
library(VisiumIO)
library(jsonlite)
library(ggspavis)
library(patchwork)
library(BiocParallel)
library(SpatialExperiment)
# specify whether/how to 
# do out parallelization
bp <- MulticoreParam(th <- 10)
# set seed for random number generation
# in order to make results reproducible
set.seed(194849)
```

## Data import

```{r load-data, message=FALSE, results="hide"}
# retrieve dataset from OSF repo
id <- "https://osf.io/5n4q3"
df <- osf_retrieve_node(id)
id <- "Visium_HumanColon_Oliveria"
fd <- osf_ls_files(df, id)
dir.create(td <- tempfile())
osf_download(fd, td, recurse=TRUE)
```

```{r spe}
# read into 'SpatialExperiment'
obj <- TENxVisium(
    spacerangerOut=td, 
    format="h5", 
    images="lowres")
(spe <- import(obj))
```

```{r plt-hne, echo=FALSE, fig.width=12, fig.height=4.5}
xy <- spatialCoords(spe)*scaleFactors(spe)
xs <- range(xy[, 1]); ys <- nrow(imgRaster(spe))-range(xy[, 2])
box <- geom_rect(
    xmin=xs[1], xmax=xs[2], ymin=ys[1], ymax=ys[2],
    col="black", fill=NA, linetype=2, linewidth=2/3)
plotVisium(spe, spots=FALSE, point_size=.75) + box +
plotVisium(spe, point_size=1.5, zoom=TRUE) +
plot_layout(nrow=1) & facet_null()
```

## Quality control

```{r add-qc}
# use gene symbols as feature names
rownames(spe) <- make.unique(rowData(spe)$Symbol)
# add per-cell quality control metrics
sub <- list(mt=grep("^MT-", rownames(spe)))
spe <- addPerCellQCMetrics(spe, subsets=sub)
```

```{r plt-xy-qc, echo=FALSE, message=FALSE, fig.width=7, fig.height=3}
spe$log_sum <- log1p(spe$sum)
plotSpots(spe, annotate="log_sum") + ggtitle("library size") +
plotSpots(spe, annotate="subsets_mt_percent") + ggtitle("% mitochondrial") +
plot_layout(nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

```{r ol}
# determine outliers via thresholding on MAD from the median
ol <- perCellQCFilters(spe, sub.fields="subsets_mt_percent")
# add results as cell metadata
colData(spe)[names(ol)] <- ol 
# tabulate # and % of cells that'd 
# be discarded for different reasons
data.frame(
    check.names=FALSE,
    `#`=apply(ol, 2, sum), 
    `%`=round(100*apply(ol, 2, mean), 2))
```

Let's see which spots would be excluded according to the above criteria:

```{r plt-ol-xy, fig.width=12, fig.height=3}
#| code-fold: true
lapply(names(ol), \(.) 
    plotSpots(spe, annotate=.) + ggtitle(.)) |>
    wrap_plots(nrow=1, guides="collect") &
    guides(col=guide_legend(override.aes=list(size=3))) &
    scale_color_manual("discard", values=c("lavender", "purple")) &
    theme(plot.title=element_text(hjust=0.5), legend.key.size=unit(0, "lines"))
```

It seems like low-quality spots are highly spatially organized, so that might might decide to refrain from removing them, for now. We will see further below how the quality control metrics used here, and spots deemed to be `discarded`, are distributed across (transcription-based) clusters.

## Processing

```{r pro}
# log-library size normalization
spe <- logNormCounts(spe)
# highly variable feature selection
tbl <- modelGeneVar(spe)
sel <- getTopHVGs(tbl, n=2e3)
# principal component analysis
spe <- runPCA(spe, subset_row=sel)
```

## Clustering

```{r clu}
# build shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(spe, use.dimred="PCA", type="jaccard")
# cluster via Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=1)
table(spe$Leiden <- factor(k$membership))
```

## Exploratory

```{r plt-clu-xy, message=FALSE, fig.width=3.5, fig.height=3}
#| code-fold: true
#| fig-align: center
plotSpots(spe, annotate="Leiden") +
    theme(legend.key.size=unit(0, "lines")) +
    scale_color_manual(values=unname(pals::trubetskoy()))
```

Let's inspect the key drivers of (expression) variability in terms of PCs:

```{r plt-pcs-xy, message=FALSE, fig.width=10, fig.height=3}
#| code-fold: true
pcs <- reducedDim(spe, "PCA")
colData(spe)[colnames(pcs)] <- pcs
lapply(colnames(pcs)[seq_len(3)], 
    \(.) plotSpots(spe, annotate=.) +
    scale_color_gradientn(., colors=pals::jet())) |>
    wrap_plots(nrow=1) & theme(
        plot.title=element_blank(),
        legend.key.width=unit(0.5, "lines"),
        legend.key.height=unit(1, "lines"))
```

Quality control metrics tend to be low for specific clusters. Their patch-like 
pattern, in turn, explains the clustering of low-quality spots seen earlier.

```{r plt-clu-qc, message=FALSE, fig.width=8, fig.height=2.5}
#| code-fold: true
lapply(c("detected", "log_sum", "subsets_mt_percent"), \(.)
    plotColData(spe, x=., y="Leiden", color_by="discard", point_size=0.1) +
    scale_x_discrete(limits=names(sort(by(spe[[.]], spe$Leiden, median))))) |>
    wrap_plots(nrow=1, guides="collect") &
    scale_color_manual(values=c("lavender", "purple")) &
    guides(col=guide_legend(override.aes=list(alpha=1, size=3))) &
    theme_minimal() & theme(
        panel.grid.minor=element_blank(), 
        legend.key.size=unit(0, "lines"))
```

## Annotation

```{r}

```

## Signatures

[Note that such gene lists like these may come from many different places - e.g., in-house analyses, other publications befitting the research question etc. As such, they may also be read into are from a *.csv* file, or stem from upstream analyses. In any case, they should be curated well and interpreted with caution.]{.aside}
Rather than investigating single genes, we can also evaluate the expression of sets of genes (e.g., pathway signatures); e.g., malignant tissue may differ in metabolic activity such as glycolysis and fatty acid metabolism, or exhibit increased apoptosis (cell death) etc. 
Here, we retrieve hallmark gene sets for said biological phenomena from [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb):

```{r MSigDB}
url <- paste0(
    "https://www.gsea-msigdb.org/gsea/msigdb/human/",
    "download_geneset.jsp?geneSetName=%s&fileType=json")
ids <- c(
    hypoxia="HYPOXIA",
    apoptosis="APOPTOSIS",
    glycolysis="GLYCOLYSIS",
    FA_metabolism="FATTY_ACID_METABOLISM")
set <- lapply(ids, \(id) {
    id <- paste0("HALLMARK_", id)
    tf <- tempfile(fileext=".json")
    download.file(sprintf(url, id), tf, quiet=TRUE)
    gs <- fromJSON(tf)[[1]]$geneSymbols
    gs <- match(gs, rowData(spe)$Symbol, nomatch=0)
    rownames(spe)[gs]
})
# we now have a list of gene sets
# (element is a character vector)
head(set$FA_metab)
# counts number of genes in each set
sapply(set, length)
```

[By definition, `AUCell` yields values in [0,1]. Larger sets (more genes) are more likely to achieve higher scores by chance (e.g., a gene set of *all* genes would score 1 in any dataset). It is thus unfair to compare them directly. However, spatial distribution, correlation, and relative comparisons between subpopulations etc. are still meaningful.]{.aside}
Next, we will score these using `r BiocStyle::Biocpkg("AUCell")` [@Aibar2017-AUCell], which works in two steps: (i) rank genes for every observation (here, spots), and (ii) compute AUC values for each gene set. In essence, these represent the fraction of genes (within top-ranked genes; default 5%) that are in a given set; i.e., high values correspond to high activity (in terms of coordinated gene expression).

```{r AUCell}
# realize (sparse) gene expression matrix
mtx <- as(logcounts(spe), "dgCMatrix") 
# build per-spot gene rankings
rnk <- AUCell_buildRankings(mtx, BPPARAM=bp, plotStats=FALSE, verbose=FALSE)
# calculate AUC for each gene set in each spot
auc <- AUCell_calcAUC(geneSets=set, rankings=rnk, nCores=th, verbose=FALSE)
# add results as spot metadata
colData(spe)[rownames(auc)] <- t(assay(auc)) 
```

```{r plt-auc-xy, message=FALSE, fig.width=11, fig.height=3}
#| code-fold: true
lapply(rownames(auc), \(.) {
    spe[[.]] <- scale(spe[[.]]) # scaling
    plotSpots(spe, annotate=.) # plotting
}) |> 
    # arrange & prettify
    wrap_plots(nrow=1, guides="collect") & 
    scale_color_gradientn(
        colors=pals::jet(),
        oob=scales::squish, 
        limits=c(-2.5, 2.5)) & 
    theme(
        legend.key.width=unit(0.5, "lines"), 
        legend.key.height=unit(1, "lines")) 
```

We can also compare (average) signature scores between subpopulations:

```{r plt-auc-hm, fig.width=6, fig.height=2}
#| code-fold: true
# aggregate AUC values by cluster
mu <- aggregateAcrossCells(auc, spe$Leiden, 
    use.assay.type="AUC", statistics="mean")
# visualize as (set x cluster) heatmap
plotHeatmap(mu, color=pals::coolwarm(),
    features=rownames(mu), exprs_values="AUC", 
    center=TRUE, scale=TRUE, show_colnames=TRUE)
```

## Appendix

### References {.unnumbered}